---
title: "Population structure"
date: "`r Sys.Date()`"
output: 
    github_document:
        toc: true
---

This directory contains population structure results. The script 
`population-structure.Rmd` renders this `README` and is responsible for creating
the relevant outputs.

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
    echo = TRUE, message = FALSE, 
    warning = FALSE, error = FALSE,
    fig.height = 10, fig.width = 10, fig.dpi=300,
    out.width = "100%"
)

suppressPackageStartupMessages({
    library(tidyverse)
    library(patchwork)
    library(here)
    library(gt)
})

theme_set(theme_bw())

pops <- fs::dir_ls(here("data", "popmaps"), glob = "*.txt") |>
    read_delim(
        delim = " ", 
        col_names = c("sample", "population"), 
        col_types = cols(), 
        id = "species"
    ) |>
    mutate(species = str_remove(basename(species), "-.*"))

ignore <- fs::dir_ls(here("results", "population-structure"), glob = "*.txt") |>
    read_tsv(
        col_names = "population"
    ) |>
    distinct() |>
    pull()

meta <- fs::dir_ls(here("data", "sample-sheets"), glob = "*csv") |>
    map(\(x) {
        tmp <- x |>
            read_csv(col_types = cols())
        if("genus" %in% colnames(tmp)) tmp <- rename(
            tmp, 
            Genus = genus,
            Species = species,
            Region = region,
            Country = country
        )
        tmp
    }) |>
    list_rbind()
```

## Helper functions

```{r helper-function-structure, echo=FALSE}
# No longer need this
readStructure <- function(path) {
    k <- as.numeric(sub(".+-(.*)", "\\1", str_remove(basename(path), "\\..*") ))
    kpops <- paste0("k", 1:k)
    read_lines(path) |>
        str_remove("^\\s+") |>
        str_remove("\\s+$") |>
        str_remove(": ") |>
        str_replace_all("\\s+", "\t") |>
        as_tibble() |>
        separate(col = value, into = c("i1", "i2", "i3", "i4", kpops), sep = "\t") |>
        select(i1, starts_with("k")) |>
        pivot_longer(names_to = "k-value", values_to = "proportion", starts_with("k")) |>
        mutate(
            proportion = as.numeric(proportion),
            i1 = as.numeric(i1)
        )
}
```

This helper function is for generating the PCAs. It handles finding `x` and `y`
maxima/minima and formatting.

```{r helper-function-plotting}
# Function that takes the PCA list we'll generate below and plot the desired PCs
# for the specified population
plotPCA <- function(pca_list, species, pcA, pcB, xstep = 1, ystep = 1, pal = "Spectral") {
    sp_pca <- pca_list |>
        pluck(species)
    
    # x-axis: max/min
    xmin <- sp_pca |>
        pull(pcA) |>
        min(na.rm = TRUE) |>
        floor()
    
    xmax <- sp_pca |>
        pull(pcA) |>
        max(na.rm = TRUE) |>
        ceiling()
    
    # y-axis: max/min
    ymin <- sp_pca |>
        pull(pcB) |>
        min(na.rm = TRUE) |>
        floor()
    
    ymax <- sp_pca |>
        pull(pcB) |>
        max(na.rm = TRUE) |>
        ceiling()
    
    sp_pca |>
        ggplot(aes(x = .data[[pcA]], y = .data[[pcB]], shape = population, colour = order, text = sample)) +
        geom_point(size = 2) +
        scale_color_brewer(palette = pal) +
        labs(title = glue::glue("{pcA} vs {pcB}")) +
        scale_y_continuous(
            limits = c(ymin, ymax),
            breaks = seq(ymin, ymax, by = ystep),
            labels = as.character(seq(ymin, ymax, by = ystep)),
            expand = c(0, 0)
        ) +
        scale_x_continuous(
            limits = c(xmin, xmax),
            breaks = seq(xmin, xmax, by = xstep),
            labels = as.character(seq(xmin, xmax, by = xstep)),
            expand = c(0, 0)
        ) +
        theme(
            legend.position = "bottom",
            legend.title = element_text(hjust = 0.5, face = "bold", size = 14),
            legend.title.position = "top",
            plot.title = element_text(hjust = 0.5),
            title = element_text(face = "bold")
        )
}
```

# Removed samples

A range of samples were excluded from the PCAs below. Further numerous populations
were ignored when running `STRUCTURE`. These can be found in the following files

- [dropped-samples.tsv][dropped]
- Per-species populations to ignore ([here][ignore])

## PCA

Principal component analysis (PCA) was conducted using the [Ipyrad analysis toolkit][ipatools].
The code responsible for the CSV files we're loading in below can be found
[here][pcaipynb]. Below, we're simply creating custom figures. We generate PCAs
for **all** samples across all populations for each species.

```{r}
lst_pca <- fs::dir_ls(here("results", "population-structure", "pca"), glob = "*.csv") |>
    # For each vector element, we're setting the basename of the file as its name
    (\(x) set_names(x, str_remove(basename(x), "-.*")))() |>
    # Read the contents of each CSV and set the column names to PC... 
    imap(\(x, i) {
        tmp <- x |>
            read_csv(col_names = TRUE, col_types = cols())
        colnames(tmp) <- c("sample", paste0("PC", 1:(ncol(tmp) - 1)))
        pops |>
            filter(species == i) |>
            left_join(tmp) |>
            left_join(meta, by = join_by(sample == id_clean))
    })
```

### *Aipysurus laevis*

Below is a patchworked plot of the first three PCs compared to each other for
*A. laevis*.

```{r pca-ALA}
pc_1_2 <- plotPCA(pca_list = lst_pca, species = "ALA", pcA = "PC1", pcB = "PC2", xstep = 2)
pc_1_3 <- plotPCA(pca_list = lst_pca, species = "ALA", pcA = "PC1", pcB = "PC3", xstep = 2)
pc_2_3 <- plotPCA(pca_list = lst_pca, species = "ALA", pcA = "PC2", pcB = "PC3")


design <- "12\n34"

pc_1_2 + pc_1_3 + pc_2_3 + guide_area() +
    plot_layout(
        design = design,
        guides = "collect",
        axes = "collect"
    ) &
    guides(colour = guide_legend(ncol = 3, override.aes = list(size = 3)))
```

### *Hydrophis major*

Below is a patchworked plot of the first three PCs compared to each other for
*H. major*.

```{r}
pc_1_2 <- plotPCA(pca_list = lst_pca, species = "HMA", pcA = "PC1", pcB = "PC2", xstep = 2, ystep = 2)
pc_1_3 <- plotPCA(pca_list = lst_pca, species = "HMA", pcA = "PC1", pcB = "PC3", xstep = 2, ystep = 2)
pc_2_3 <- plotPCA(pca_list = lst_pca, species = "HMA", pcA = "PC2", pcB = "PC3", ystep = 2)

pc_1_2 + pc_1_3 + pc_2_3 + guide_area() +
    plot_layout(
        design = design, 
        guides = "collect",
        axes = "collect"
    ) &
    guides(colour = guide_legend(ncol = 3, override.aes = list(size = 4)))
```

### *Hydrophis stokesii*

Below is a patchworked plot of the first three PCs compared to each other for
*H. stokesii*.

```{r}
pc_1_2 <- plotPCA(pca_list = lst_pca, species = "HST", pcA = "PC1", pcB = "PC2", xstep = 2, ystep = 2)
pc_1_3 <- plotPCA(pca_list = lst_pca, species = "HST", pcA = "PC1", pcB = "PC3", xstep = 2, ystep = 2)
pc_2_3 <- plotPCA(pca_list = lst_pca, species = "HST", pcA = "PC2", pcB = "PC3", xstep = 2, ystep = 2)

pc_1_2 + pc_1_3 + pc_2_3 + guide_area() +
    plot_layout(
        design = design, 
        guides = "collect",
        axes = "collect"
    ) &
    guides(colour = guide_legend(ncol = 3, override.aes = list(size = 4)))
```

## STRUCTURE

`STRUCTURE` plots are then generated for WA populations, specifically: Broome,
Pilbara, Exmouth Gulf, Shark Bay and North Kimberley. In the PCAs above, samples
from these populations typically cluster together, showing little structure.
Consequently, we're using this method as a more granular approach to determine
if there is any allele sharing between the populations.

```{r read-structure}
structure_pops <- unique(pops$population)[ ! unique(pops$population) %in% ignore ]

# Custom ordering of colour palette
pal <- c(
    "#8da0cb", "#66c2a5", "#fc8d62", 
    "#e78ac3", "#a6d854", "#ffd92f", 
    "#f0ad4e", "#d9534f", "#5cb85c"
)
names(pal) <- 1:9

structure_df <- fs::dir_ls(
    here("results", "population-structure", "structure"), 
    glob = "*plot.csv", 
    recurse = TRUE
) |>
    map(\(x) {
        tmp <- x |>
            read_csv(col_names = TRUE, col_types = cols())
        tmp |>
            mutate(K = glue::glue("K{ncol(tmp) - 1}")) |>
            pivot_longer(names_to = "partition", values_to = "proportion", 2:ncol(tmp)) |>
            mutate(partition = as.numeric(partition) + 1) |>
            rename(sample = `...1`) |>
            left_join(pops)
    }) |>
    list_rbind() |>
    mutate(
        partition = forcats::fct_inorder(as.character(partition)), 
        population = factor(population, levels = structure_pops),
        .by = K
    )
```

### Choosing the correct K (N-ancestral populations)

Choosing the best `K` value can be determined by plotting the estimated
log-probability for each `K` value, in addition to the delta-K value. The values
can be interpreted as follows:

- **Estimated log-probability**: Higher values equal better model fit.
- **Delta-K**: The rate of change between successive K-values. Larger values
indicate greater change in the model fit.

The results for the three species are shown below.

```{r}
etable <- fs::dir_ls(
    here("results", "population-structure", "structure"), 
    glob = "*etable.csv",
    recurse = TRUE
) |>
    read_csv(col_names = TRUE, col_types = cols(), id = "species") |>
    mutate(species = str_remove(basename(species), "-etable.csv")) |>
    rename(K = `...1`)

etable |>
    gt(groupname_col = "species") |>
    fmt_number(
        columns = c("K", "Nreps"),
        drop_trailing_zeros = TRUE
    ) |>
    fmt_number(
        columns = 3:7,
        n_sigfig = 4
    ) |>
    as_raw_html()
```

Plotting the table above looks like the following.

```{r}
prb <- etable |>
    ggplot(aes(x = K, y = estLnProbMean, colour = species)) +
    geom_point() +
    geom_line() +
    scale_y_continuous(
        name = "Estimated mean log-probability",
        breaks = seq(0, -105000, -10000),
        limits = c(-105e3, -3e4),
        labels = scales::label_number(style_negative = "minus")
    ) +
    scale_x_continuous(breaks = seq(2, 9, 1))

dk <- etable |>
    ggplot(aes(x = K, y = deltaK, colour = species)) +
    geom_point() +
    geom_line() +
    scale_x_continuous(breaks = seq(2, 9, 1)) +
    scale_y_continuous(
        name = "Delta-K",
        breaks = seq(0, 15, 1)
    )

prb/dk +
    plot_layout(axes = "collect", guides = "collect")
```

Estimated mean log probabilities are relatively consistent across the first four
to six K-values in each population, before starting to dip. Delta-K values vary
quite a bit for *H. major*, in the lower K-values, while *A. laevis* and *H. stokesii*
peak at `K = 7`. The issue is that the estimated log probabilities don't really
improve that much for the increasing K, meaning the delta-K peaks are likely a
product of over-fitting.

### *Aipysurus laevis*

The figure below plots all tested population k-values (rows). As is expected
based on the log-probability + delta-k figure above, adding populations doesn't
really seem to do much, as the additional populations have minute contributions
overall.

As such, a conservative approach using a lower K-value (i.e. fewer populations)
is likely the play.

```{r}
structure_df |>
    filter(species == "ALA") |>
    ggplot(aes(x = sample, y = proportion, fill = partition)) +
    geom_col(position = "stack") +
    scale_fill_manual(values = pal) +
    scale_y_continuous(
        limits = c(0, 1),
        breaks = c(0, 0.5, 1),
        labels = as.character(seq(0, 1, 0.5)),
        expand = c(0, 0)
    ) +
    labs(x = NULL) + 
    facet_grid(
        rows = vars(K), cols = vars(population),
        scales = "free_x",
        space = "free_x"
    ) +
    guides(fill = guide_legend(title.position = "bottom")) +
    theme(
        axis.title.y = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.98),
        legend.position = "bottom",
        legend.title = element_text(hjust = 0.5),
        strip.text.x = element_text(
            face = "bold", 
            hjust = 0.5,
            size = 14
        ),
        strip.text.y = element_text(size = 14, hjust = 0.5),
        strip.background = element_blank(),
        panel.spacing.x = unit(0.2, "line"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()
    )
```

Below is the figure for `K = 2` and `K = 3`. Even adding an additional population
(`K = 3`) seems to be over-fitting.

```{r}
structure_df |>
    filter(species == "ALA", K %in% c("K2", "K3")) |>
    ggplot(aes(x = sample, y = proportion, fill = partition)) +
    geom_col(position = "stack") +
    scale_fill_manual(values = pal) +
    scale_y_continuous(
        limits = c(0, 1),
        breaks = c(0, 0.5, 1),
        labels = as.character(seq(0, 1, 0.5)),
        expand = c(0, 0)
    ) +
    labs(x = NULL) + 
    facet_grid(
        rows = vars(K), cols = vars(population),
        scales = "free_x",
        space = "free_x"
    ) +
    guides(fill = guide_legend(title.position = "bottom")) +
    theme(
        axis.title.y = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.98),
        legend.position = "bottom",
        legend.title = element_text(hjust = 0.5),
        strip.text.x = element_text(
            face = "bold", 
            hjust = 0.5,
            size = 14
        ),
        strip.text.y = element_text(size = 14, hjust = 0.5),
        strip.background = element_blank(),
        panel.spacing.x = unit(0.2, "line"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()
    )
```

### *Hydrophis major*

Here, we do the same for *H. major*.

```{r}
structure_df |>
    filter(species == "HMA") |> 
    ggplot(aes(x = sample, y = proportion, fill = partition)) +
    geom_col(position = "stack") +
    scale_fill_manual(values = pal) +
    scale_y_continuous(
        breaks = c(0, 0.5, 1),
        labels = as.character(seq(0, 1, 0.5)),
        expand = c(0, 0)
    ) +
    labs(x = NULL) + 
    facet_grid(
        rows = vars(K), cols = vars(population),
        scales = "free_x",
        space = "free_x"
    ) +
    guides(fill = guide_legend(title.position = "bottom")) +
    theme(
        axis.title.y = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.98),
        legend.position = "bottom",
        legend.title = element_text(hjust = 0.5),
        strip.text.x = element_text(
            face = "bold", 
            hjust = 0.5,
            size = 14
        ),
        strip.text.y = element_text(size = 14, hjust = 0.5),
        strip.background = element_blank(),
        panel.spacing.x = unit(0.2, "line"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()
    )
```

And again plot the first few K-increments. There does look to be a bit of
variability in the Shark Bay samples when `K = 3`, so perhaps a higher K-value
is warranted. However, K-values of 4 and 5 appear to be over fitting.

```{r}
structure_df |>
    filter(species == "HMA", K %in% c("K2", "K3", "K4", "K5")) |>
    ggplot(aes(x = sample, y = proportion, fill = partition)) +
    geom_col(position = "stack") +
    scale_fill_manual(values = pal) +
    scale_y_continuous(
        breaks = c(0, 0.5, 1),
        labels = as.character(seq(0, 1, 0.5)),
        expand = c(0, 0)
    ) +
    labs(x = NULL) + 
    facet_grid(
        rows = vars(K), cols = vars(population),
        scales = "free_x",
        space = "free_x"
    ) +
    guides(fill = guide_legend(title.position = "bottom")) +
    theme(
        axis.title.y = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.98),
        legend.position = "bottom",
        legend.title = element_text(hjust = 0.5),
        strip.text.x = element_text(
            face = "bold", 
            hjust = 0.5,
            size = 14
        ),
        strip.text.y = element_text(size = 14, hjust = 0.5),
        strip.background = element_blank(),
        panel.spacing.x = unit(0.2, "line"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()
    )
```

### *Hydrophis stokesii*

Lastly, we do the same for *H. stokesii*.

```{r}
structure_df |>
    filter(species == "HST") |> 
    ggplot(aes(x = sample, y = proportion, fill = partition)) +
    geom_col(position = "stack") +
    scale_fill_manual(values = pal) +
    scale_y_continuous(
        breaks = c(0, 0.5, 1),
        labels = as.character(seq(0, 1, 0.5)),
        expand = c(0, 0)
    ) +
    labs(x = NULL) + 
    facet_grid(
        rows = vars(K), cols = vars(population),
        scales = "free_x",
        space = "free_x"
    ) +
    guides(fill = guide_legend(title.position = "bottom")) +
    theme(
        axis.title.y = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.98),
        legend.position = "bottom",
        legend.title = element_text(hjust = 0.5),
        strip.text.x = element_text(
            face = "bold", 
            hjust = 0.1,
            vjust = 0.5,
            size = 14,
            angle = 90
        ),
        strip.text.y = element_text(size = 14, hjust = 0.5),
        strip.background = element_blank(),
        panel.spacing.x = unit(0.2, "line"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()
    )
```

Similar to the other species, these snakes appear to be a homogeneous population
in the WA region, as `K = 3` appears to simply be fractioning the starting populations.

```{r}
structure_df |>
    filter(species == "HST", K %in% c("K2", "K3")) |>
    ggplot(aes(x = sample, y = proportion, fill = partition)) +
    geom_col(position = "stack") +
    scale_fill_manual(values = pal) +
    scale_y_continuous(
        breaks = c(0, 0.5, 1),
        labels = as.character(seq(0, 1, 0.5)),
        expand = c(0, 0)
    ) +
    labs(x = NULL) + 
    facet_grid(
        rows = vars(K), cols = vars(population),
        scales = "free_x",
        space = "free_x"
    ) +
    guides(fill = guide_legend(title.position = "bottom")) +
    theme(
        axis.title.y = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.98),
        legend.position = "bottom",
        legend.title = element_text(hjust = 0.5),
        strip.text.x = element_text(
            face = "bold", 
            hjust = 0.1,
            vjust = 0.5,
            size = 14,
            angle = 90
        ),
        strip.text.y = element_text(size = 14, hjust = 0.5),
        strip.background = element_blank(),
        panel.spacing.x = unit(0.2, "line"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()
    )
```

[ipatools]: https://ipyrad.readthedocs.io/en/master/API-analysis/index.html
[pcaipynb]: https://github.com/a-lud/sea-snake-dart/blob/main/scripts/06-pca.ipynb
[dropped]: 
[ignore]: 
