---
title: "Population structure"
date: "`r Sys.Date()`"
output: 
    github_document:
        toc: true
---

This directory contains population structure results. The script 
`population-structure.Rmd` renders this `README` and is responsible for creating
the relevant outputs.

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
    echo = TRUE, message = FALSE, 
    warning = FALSE, error = FALSE,
    fig.height = 10, fig.width = 10, fig.dpi=300,
    out.width = "100%"
)

suppressPackageStartupMessages({
    library(tidyverse)
    library(patchwork)
    library(here)
    library(gt)
})

theme_set(theme_bw())

pops <- fs::dir_ls(here("data", "popmaps"), glob = "*.txt") |>
    read_delim(
        delim = " ", 
        col_names = c("sample", "population"), 
        col_types = cols(), 
        id = "species"
    ) |>
    mutate(
        population = str_replace_all(population, "_", " "),
        species = str_remove(basename(species), "-.*")
    )

ignore <- fs::dir_ls(here("results", "population-structure"), glob = "*.txt") |>
    read_tsv(
        col_names = "population",
        col_types = cols()
    ) |>
    distinct() |>
    pull() |>
    str_replace_all("_", " ")

# Problem samples to remove 
samples_remove <- read_tsv(
    here("data", "popmaps", "dropped-samples.tsv"), 
    skip = 1, 
    col_names = c("sample", "pop"),
    col_types = cols()
) |>
    pull(sample)
```

## Helper functions

This helper function is for generating the PCAs. It handles finding `x` and `y`
maxima/minima and formatting.

```{r helper-function-plotting}
# Function that takes the PCA list we'll generate below and plot the desired PCs
# for the specified population
plotPCA <- function(pca_dataframe, variance_dataframe, spec, WA_only, pcA, pcB, xstep = 1, ystep = 1, pal) {
    full_species <- set_names(c("ALA", "HMA", "HST"), c("Aipysurus laevis", "Hydrophis major", "Hydrophis stokesii"))
    plot_title <- names(full_species[full_species == spec])
    
    sp_pca <- pca_dataframe |>
        filter(species == spec, WA == WA_only)
    
    sp_var <- variance_dataframe |>
        filter(species == spec, WA == WA_only)
    
    # TODO: Figure out why this is being a piece of shit
    if(WA_only){
        wa_levels <- pca_dataframe |> 
            filter(WA == TRUE) |> 
            pull(population) |> 
            unique() |>
            as.character()
        
        sp_pca <- sp_pca |>
            mutate(
                population = factor(
                    population,
                    levels = wa_levels
                )
            )
    }
    
    # x-axis: max/min
    xmin <- sp_pca |>
        pull(pcA) |>
        min(na.rm = TRUE) |>
        floor()
    
    xmax <- sp_pca |>
        pull(pcA) |>
        max(na.rm = TRUE) |>
        ceiling()
    
    # y-axis: max/min
    ymin <- sp_pca |>
        pull(pcB) |>
        min(na.rm = TRUE) |>
        floor()
    
    ymax <- sp_pca |>
        pull(pcB) |>
        max(na.rm = TRUE) |>
        ceiling()
    
    # Variance strings
    pcA_var <- round(
        sp_var |>
            filter(pc_axes == pcA) |>
            pull(variance) * 100, digits = 2
    )
    
    pcB_var <- round(
        sp_var |>
            filter(pc_axes == pcB) |>
            pull(variance) * 100, digits = 2
    )
    
    sp_pca |>
        ggplot(
            aes(
                x = .data[[pcA]], y = .data[[pcB]],
                colour = population
            )
        ) +
        geom_point(size = 2.5, show.legend=TRUE) +
        scale_colour_manual(
            values = pal,
            drop = FALSE
        ) +
        labs(
            title = glue::glue("{plot_title}"),
            colour = "Population",
            x = glue::glue("{pcA} ({pcA_var}%)"),
            y = glue::glue("{pcB} ({pcB_var}%)")
        ) +
        scale_y_continuous(
            limits = c(ymin, ymax),
            breaks = seq(ymin, ymax, by = ystep),
            labels = as.character(seq(ymin, ymax, by = ystep)),
            expand = c(0, 0)
        ) +
        scale_x_continuous(
            limits = c(xmin, xmax),
            breaks = seq(xmin, xmax, by = xstep),
            labels = as.character(seq(xmin, xmax, by = xstep)),
            expand = c(0, 0)
        ) +
        theme(
            axis.text = element_text(size = 14),
            axis.title = element_text(size = 14),
            legend.position = "bottom",
            legend.title = element_text(hjust = 0.5, size = 14),
            legend.title.position = "top",
            legend.text = element_text(size = 14),
            plot.title = element_text(hjust = 0.5, face = "bold.italic")
        )
}
```

# Removed samples

A range of samples were excluded from the PCAs below. Further numerous populations
were ignored when running `STRUCTURE`. These can be found in the following files

- [dropped-samples.tsv][dropped]
- Per-species populations to ignore ([here][ignore])

## PCA

Principal component analysis (PCA) was conducted using the [Ipyrad analysis toolkit][ipatools].
The code responsible for the CSV files we're loading in below can be found
[here][pcaipynb]. Below, we're simply creating custom figures. We generate PCAs
for **all** samples across all populations for each species.

```{r load-pca-data}
pop_levels <- c(
    "Shark Bay" ,"Exmouth Gulf", "Pilbara", 
    "Broome", "North Kimberley", "Scott Reef", "Ashmore", "Heywood Shoal",
    "Gulf of Carpentaria", "North QLD", "South QLD", "New Caledonia"
)

# PCA dataframes
df_pca <- fs::dir_ls(here("results", "population-structure", "pca"), glob = "*.csv") |>
    (\(x) x[!str_detect(x, "variance")])() |>
    imap(\(x, i) {
        fname <- str_remove(basename(i), ".csv")
        spec <- str_remove(fname, "-.*")
        wa <- if(str_detect(fname, "WA")) TRUE else FALSE
        
        tmp <- x |>
            read_csv(col_names = TRUE, col_types = cols())
        colnames(tmp) <- c("sample", paste0("PC", 1:(ncol(tmp) - 1)))
        pops |>
            filter(species == spec) |>
            inner_join(tmp) |>
            mutate(WA = wa, .after = species)
    }) |>
    list_rbind() |>
    select(species, WA, sample, population, PC1, PC2, PC3) |>
    mutate(
        population = str_replace_all(population, "_", " "),
        population = factor(population, levels = pop_levels)
    )

# Variance dataframes
df_variance <- fs::dir_ls(here("results", "population-structure", "pca"), glob = "*variance.csv") |>
    imap(\(x, i) {
        fname <- str_remove(basename(i), ".csv")
        spec <- str_remove(fname, "-.*")
        wa <- if(str_detect(fname, "WA")) TRUE else FALSE
        x |>
            read_csv(
                skip = 1,
                col_names = c("pc_axes", "variance"),
                col_types = cols()
            ) |>
            mutate(
                species = spec,
                WA = wa,
                pc_axes = pc_axes + 1,
                pc_axes = paste0("PC", pc_axes)
            ) |>
            filter(pc_axes %in% c("PC1", "PC2", "PC3"))
    }) |>
    list_rbind()
```

```{r get-legend}
# Custom ordering of colour palette
pal_PCA <- c(
    "#8da0cb", "#66c2a5", "#fc8d62", 
    "#e78ac3", "#a6d854", "#ffd92f", 
    "#f0ad4e", "#d9534f", "#5cb85c",
    "#66C2A5", "#FC8D62", "#8DA0CB",
    "#B3B3B3"
)
names(pal_PCA) <- pop_levels
```

### All samples

```{r pca-ALA}
ala_pc_1_2 <- plotPCA(pca_dataframe = df_pca, variance_dataframe = df_variance, WA_only = FALSE, spec = "ALA", pcA = "PC1", pcB = "PC2", xstep = 2, pal = pal_PCA)
hst_pc_1_2 <- plotPCA(pca_dataframe = df_pca, variance_dataframe = df_variance, WA_only = FALSE, spec = "HMA", pcA = "PC1", pcB = "PC2", xstep = 2, ystep = 2, pal = pal_PCA)
hma_pc_1_2 <- plotPCA(pca_dataframe = df_pca, variance_dataframe = df_variance, WA_only = FALSE, spec = "HST", pcA = "PC1", pcB = "PC2", xstep = 2, ystep = 2, pal = pal_PCA)

design <- "12\n34"

plot_pca_all <- ala_pc_1_2 + hst_pc_1_2 + hma_pc_1_2 + guide_area() +
    plot_layout(
        design = design,
        guides = "collect",
        axes = "collect"
    ) +
    plot_annotation(tag_levels = 'A') &
    guides(colour = guide_legend(ncol = 2, override.aes = list(size = 3))) &
    theme(plot.tag = element_text(size = 14, face = "bold"))

ragg::agg_png(
    filename = here("results", "population-structure", "pca-all-species-all-pops.png"),
    width = 1500, 
    height = 1500,
    units = "px",  
    res = 150
)
plot_pca_all
invisible(dev.off())

plot_pca_all
```

### WA Coast

And the same for *just* the WA coastline samples

```{r}
ala_pc_1_2 <- plotPCA(pca_dataframe = df_pca, variance_dataframe = df_variance, WA_only = TRUE, spec = "ALA", pcA = "PC1", pcB = "PC2", xstep = 2, ystep = 2, pal = pal_PCA)
hma_pc_1_2 <- plotPCA(pca_dataframe = df_pca, variance_dataframe = df_variance, WA_only = TRUE, spec = "HMA", pcA = "PC1", pcB = "PC2", xstep = 2, ystep = 2, pal = pal_PCA)
hst_pc_1_2 <- plotPCA(pca_dataframe = df_pca, variance_dataframe = df_variance, WA_only = TRUE, spec = "HST", pcA = "PC1", pcB = "PC2", xstep = 2, ystep = 2, pal = pal_PCA)

design <- "12\n34"

plot_pca_wa <- ala_pc_1_2 + hma_pc_1_2 + hst_pc_1_2 + guide_area() +
    plot_layout(
        design = design,
        guides = "collect",
        axes = "collect"
    ) &
    plot_annotation(tag_levels = 'A') &
    guides(colour = guide_legend(ncol = 2, override.aes = list(size = 3))) &
    theme(plot.tag = element_text(size = 14, face = "bold"))

ragg::agg_png(
    filename = here("results", "population-structure", "pca-all-species-WA-pops.png"),
    width = 1500, 
    height = 1500,
    units = "px",  
    res = 150
)
plot_pca_wa
invisible(dev.off())

plot_pca_wa
```

We can see that the Shark Bay samples partition off in both *A. laevis* and
*H. major*.

## STRUCTURE

`STRUCTURE` plots are then generated for WA populations, specifically: Broome,
Pilbara, Exmouth Gulf, Shark Bay and North Kimberley. In the PCAs above, samples
from these populations typically cluster together, showing little structure.
Consequently, we're using this method as a more granular approach to determine
if there is any allele sharing between the populations.

```{r read-structure}
structure_pops <- unique(pops$population)[ ! unique(pops$population) %in% ignore ]

# Custom ordering of colour palette
pal <- c(
    "#8da0cb", "#66c2a5", "#fc8d62", 
    "#e78ac3", "#a6d854", "#ffd92f", 
    "#f0ad4e", "#d9534f", "#5cb85c"
)
names(pal) <- 1:9

structure_df <- fs::dir_ls(
    here("results", "population-structure", "structure"), 
    glob = "*plot.csv", 
    recurse = TRUE
) |>
    map(\(x) {
        tmp <- x |>
            read_csv(col_names = TRUE, col_types = cols())
        tmp |>
            mutate(K = glue::glue("K{ncol(tmp) - 1}")) |>
            pivot_longer(names_to = "partition", values_to = "proportion", 2:ncol(tmp)) |>
            mutate(partition = as.numeric(partition) + 1) |>
            rename(sample = `...1`) |>
            left_join(pops)
    }) |>
    list_rbind() |>
    mutate(
        partition = forcats::fct_inorder(as.character(partition)),
        population = str_replace_all(population, "_", " "),
        population = factor(population, levels = structure_pops),
        .by = K
    )
```

### Choosing the correct K (N-ancestral populations)

Choosing the best `K` value can be determined by plotting the estimated
log-probability for each `K` value, in addition to the delta-K value. The values
can be interpreted as follows:

- **Estimated log-probability**: Higher values equal better model fit.
- **Delta-K**: The rate of change between successive K-values. Larger values
indicate greater change in the model fit.

The results for the three species are shown below.

```{r}
etable <- fs::dir_ls(
    here("results", "population-structure", "structure"), 
    glob = "*etable.csv",
    recurse = TRUE
) |>
    read_csv(col_names = TRUE, col_types = cols(), id = "species") |>
    mutate(species = str_remove(basename(species), "-.*")) |>
    rename(K = `...1`)

etable |>
    gt(groupname_col = "species") |>
    fmt_number(
        columns = c("K", "Nreps"),
        drop_trailing_zeros = TRUE
    ) |>
    fmt_number(
        columns = 3:7,
        n_sigfig = 4
    ) |>
    as_raw_html()
```

Plotting the table above looks like the following.

```{r}
prb <- etable |>
    ggplot(aes(x = K, y = estLnProbMean, colour = species)) +
    geom_point() +
    geom_line() +
    scale_y_continuous(
        name = "Estimated mean log-probability",
        breaks = seq(0, -105000, -10000),
        limits = c(-105e3, -3e4),
        labels = scales::label_number(style_negative = "minus")
    ) +
    scale_x_continuous(breaks = seq(2, 9, 1))

dk <- etable |>
    ggplot(aes(x = K, y = deltaK, colour = species)) +
    geom_point() +
    geom_line() +
    scale_x_continuous(breaks = seq(2, 9, 1)) +
    scale_y_continuous(
        name = "Delta-K",
        breaks = seq(0, 15, 1)
    )

plot_models <- prb/dk +
    plot_layout(axes = "collect", guides = "collect")

ragg::agg_png(
    filename = here("results", "population-structure", "structure-model-fit.png"),
    width = 1000, 
    height = 800,
    units = "px",  
    res = 150
)
plot_models
invisible(dev.off())

plot_models
```

Estimated mean log probabilities are relatively consistent across the first four
to six K-values in each population, before starting to dip. Delta-K values vary
quite a bit for *H. major*, in the lower K-values, while *A. laevis* and *H. stokesii*
peak at `K = 7`. The issue is that the estimated log probabilities don't really
improve that much for the increasing K, meaning the delta-K peaks are likely a
product of over-fitting.

### *Aipysurus laevis*

Below are the figures for two an three estimated populations.

```{r}
plot_structure_ALA <- structure_df |>
    filter(species == "ALA", K %in% c("K2", "K3")) |>
    ggplot(aes(x = sample, y = proportion, fill = partition)) +
    geom_col(position = "stack") +
    scale_fill_manual(values = pal) +
    scale_y_continuous(
        limits = c(0, 1),
        breaks = c(0, 0.5, 1),
        labels = as.character(seq(0, 1, 0.5)),
        expand = c(0, 0)
    ) +
    labs(x = NULL) + 
    facet_grid(
        rows = vars(K), cols = vars(population),
        scales = "free_x",
        space = "free_x"
    ) +
    guides(fill = guide_legend(title.position = "bottom")) +
    theme(
        axis.title.y = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.98),
        legend.position = "bottom",
        legend.title = element_text(hjust = 0.5),
        strip.text.x = element_text(
            face = "bold", 
            hjust = 0.5,
            size = 14
        ),
        strip.text.y = element_text(size = 14, hjust = 0.5),
        strip.background = element_blank(),
        panel.spacing.x = unit(0.2, "line"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()
    )

ragg::agg_png(
    filename = here("results", "population-structure", "structure-ALA-WA.png"),
    width = 1500, 
    height = 1000,
    units = "px",  
    res = 150
)
plot_structure_ALA
invisible(dev.off())

plot_structure_ALA
```

### *Hydrophis major*

The figure for *H. major* shows the Shark Bay samples as having a distinct
genetic profile at $\text{K}=\text{3}$.

```{r}
plot_structure_HMA <- structure_df |>
    filter(species == "HMA", K %in% c("K2", "K3")) |>
    ggplot(aes(x = sample, y = proportion, fill = partition)) +
    geom_col(position = "stack") +
    scale_fill_manual(values = pal) +
    scale_y_continuous(
        breaks = c(0, 0.5, 1),
        labels = as.character(seq(0, 1, 0.5)),
        expand = c(0, 0)
    ) +
    labs(x = NULL) + 
    facet_grid(
        rows = vars(K), cols = vars(population),
        scales = "free_x",
        space = "free_x"
    ) +
    guides(fill = guide_legend(title.position = "bottom")) +
    theme(
        axis.title.y = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.98),
        legend.position = "bottom",
        legend.title = element_text(hjust = 0.5),
        strip.text.x = element_text(
            face = "bold", 
            hjust = 0.5,
            size = 14
        ),
        strip.text.y = element_text(size = 14, hjust = 0.5),
        strip.background = element_blank(),
        panel.spacing.x = unit(0.2, "line"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()
    )

ragg::agg_png(
    filename = here("results", "population-structure", "structure-HMA-WA.png"),
    width = 1500, 
    height = 1000,
    units = "px",  
    res = 150
)
plot_structure_HMA
invisible(dev.off())

plot_structure_HMA
```

### *Hydrophis stokesii*

*H. stokesii* has little evidence to suggest more than two populations in WA.
The Delta-K figure above indicates `K=4`, but the mean log-probability doesn't
indicate that the `K=4` model is better than any of others, perhaps suggesting
over-fitting.

```{r}
plot_structure_HST <- structure_df |>
    filter(species == "HST", K %in% c("K2", "K3" )) |> 
    ggplot(aes(x = sample, y = proportion, fill = partition)) +
    geom_col(position = "stack") +
    scale_fill_manual(values = pal) +
    scale_y_continuous(
        breaks = c(0, 0.5, 1),
        labels = as.character(seq(0, 1, 0.5)),
        expand = c(0, 0)
    ) +
    labs(x = NULL) + 
    facet_grid(
        rows = vars(K), cols = vars(population),
        scales = "free_x",
        space = "free_x"
    ) +
    guides(fill = guide_legend(title.position = "bottom")) +
    theme(
        axis.title.y = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.98),
        legend.position = "bottom",
        legend.title = element_text(hjust = 0.5),
        strip.text.x = element_text(
            face = "bold", 
            hjust = 0.1,
            vjust = 0.5,
            size = 14,
            angle = 90
        ),
        strip.text.y = element_text(size = 14, hjust = 0.5),
        strip.background = element_blank(),
        panel.spacing.x = unit(0.2, "line"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()
    )

ragg::agg_png(
    filename = here("results", "population-structure", "structure-HST-WA.png"),
    width = 1500, 
    height = 1000,
    units = "px",  
    res = 150
)
plot_structure_HST
invisible(dev.off())

plot_structure_HST
```

# Isolation by Distance (IBD)

Lastly, we'll perform Isolation-by-distance analysis. The function
`calc_isolation_by_distance()` from the `snpR` package doesn't like working with
single samples as the object returned after checking if there is only one sample
is a dataframe when it should be a vector. Consequently, the matrix/array
ends up being a single row with multiple columns rather than two columns (long/lat)
and each sample as rows.

As such, I've edited the function and run the custom version below.

Note: I also found this repo helpful - https://github.com/jdalapicolla/IBD_models.R

```{r}
# Load the snpR library and the custom scripts (found in the scripts directory)
library(snpR)
source(here("scripts", "snpR-edited.R"))

wa_pops <- c("Shark_Bay", "Exmouth_Gulf", "Pilbara", "Broome", "North_Kimberley")

# Load updated lat/long popmaps
set.seed(123)
meta_latLong <- fs::dir_ls(here("data", "IBD-popmaps"), glob = "*avg.csv") |>
    (\(x) set_names(x, str_remove(basename(x), "-.*")))() |> 
    map(\(x) {
        x |>
            read_csv(col_types = cols()) |>
            select(species, id, targetid, pop, latitude, longitude) |>
            unite(col = "sampID", sep = "-", species, id, targetid) |> 
            filter(!is.na(latitude), pop %in% wa_pops, ! sampID %in% samples_remove) |>
            select(sampID, pop, x = longitude, y = latitude) |>
            # For snakes with identical long/lat + population, choose one at random
            slice_sample(n = 1, by = c(pop, x, y))
    })

# Load the VCF files as snpR data objects
vcfs <- fs::dir_ls(
    here("results", "ipyrad"),
    glob = "*stringent.vcf.gz",
    recurse = TRUE
) |>
    (\(x) set_names(x, str_remove(basename(x), "-stringent.*")))() |>
    imap(\(v,i) {
        tmp_vcf <- read_vcf(file = v)
        
        filt_vcf <- filter_snps(
            tmp_vcf,
            min_ind = 0.8,
            min_loci = 0.75,
            re_run = TRUE
        )
        
        # Subset for samples to keep + add in metadata
        filt_vcf <- subset_snpR_data(
            x = filt_vcf, 
            .facets = "sampID",
            .subfacets = meta_latLong[[i]]$sampID
        )
        
        sample.meta(filt_vcf) <- left_join(sample.meta(filt_vcf), meta_latLong[[i]])
        
        filt_vcf <- custom_calc_genetic_distances(filt_vcf, facets = c("sampID", "pop"), method = "Nei")
        filt_vcf
    })
```

After loading the VCF files and performing the mantel test, I generate a summary
table of the statistics.

```{r IBD-mantel}
# First, get statistics
mantel_statistics <- vcfs |>
    imap(\(x, i) {
        tmp <- custom_ibd(
            x = x,
            facets = c("sampID", "pop"),
            genetic_distance_method = "Nei"
        )
        
        # Extract statistic results: TODO
        tmp_stats <- get.snpR.stats(tmp, facets = "sampID", stats = "ibd")
        tmp_stats <- tmp_stats$sampID$.base$Nei
        
        tibble(
            "Observed correlation" = tmp_stats$obs,
            "pvalue" = tmp_stats$pvalue,
            "Std. Obs" = tmp_stats$expvar[[1]],
            "Expectation" = tmp_stats$expvar[[2]],
            "Variance" = tmp_stats$expvar[[3]],
            "Replicates" = tmp_stats$rep
        )
    }) |>
    list_rbind(names_to = "species")

# Write to file
mantel_statistics |>
    write_csv(here("results", "population-structure", "mantel-statistics.csv"))

mantel_statistics |>
    gt(rowname_col = "species") |>
    gt::fmt_number(columns = c(2,3,4,5,6), n_sigfig = 2) |>
    as_raw_html()
```

Lastly, let's generate a scatter plot of distance vs genetic distance. Add the
correlation coefficients and p-values in Inkscape or something.

```{r}
df_long <- vcfs |>
    imap(\(x, i) {
        tmp <- custom_ibd(
            x = x,
            facets = c("sampID", "pop"),
            genetic_distance_method = "Nei"
        )
        
        # Dataframe for plotting
        mta <- get.snpR.stats(tmp, facets = "sampID", "geo_dist")
        mta <- as.matrix(mta$sampID$Nei)
        mta[upper.tri(mta, diag = T)] = NA
        geo_dist <- mta |>
            reshape2::melt() |>
            na.omit() |>
            arrange(Var1) |>
            setNames(c("X1", "X2","GeoDist")) |>
            as_tibble()
        
        mta <- get.snpR.stats(tmp, facets = "sampID", "genetic_distance")
        mta <- as.matrix(mta$sampID$.base$Nei)
        mta[upper.tri(mta, diag = T)] = NA
        gen_dist <- mta |>
            reshape2::melt() |>
            na.omit() |>
            arrange(Var1) |>
            setNames(c("X1", "X2", "GenDist")) |>
            as_tibble()
        
        left_join(gen_dist, geo_dist)
        
    }) |>
    list_rbind(names_to = "species")

# Plot distance (kilometres) vs genetic distance (Nei)
plot_scatter <- df_long |>
    ggplot(aes(x = GeoDist, y = GenDist, colour = species)) +
    geom_point(alpha = 0.4) +
    geom_smooth(mapping = aes(colour = species), method = "lm", show.legend = FALSE) +
    scale_x_continuous(
        name = "Distance (Km)",
        breaks = seq(0, 1600, 200),
        labels = as.character(seq(0, 1600, 200)),
        expand = c(0.01, 0, 0.02, 0)
    ) +
    scale_y_continuous(
        name = "Genetic distance (Nei)",
        breaks = seq(0, 0.055, 0.005),
        labels = as.character(seq(0, 0.055, 0.005)),
        limits = c(0, 0.055),
        expand = c(0.02, 0)
    ) +
    scale_colour_brewer(palette = "Set2") +
    guides(colour = guide_legend(override.aes = list(size = 6))) +
    theme(
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        legend.position = "top",
        legend.text = element_text(size = 14),
        legend.title = element_blank()
    )

cairo_pdf(
    here("results", "population-structure", "mantel-scatter.pdf"),
    width = 12, height = 12
)
plot_scatter
invisible(dev.off())

plot_scatter
``` 

