---
title: "Population structure"
date: "`r Sys.Date()`"
output: 
    github_document:
        toc: true
---

This directory contains population structure results. The script 
`population-structure.Rmd` renders this `README` and is responsible for creating
the relevant outputs.

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
    echo = TRUE, message = FALSE, 
    warning = FALSE, error = FALSE,
    fig.height = 10, fig.width = 10, fig.dpi=300,
    out.width = "100%"
)

suppressPackageStartupMessages({
    library(paletteer)
    library(tidyverse)
    library(patchwork)
    library(here)
    library(gt)
})

theme_set(theme_bw())

# Set the populations
wa_pops <- c("Shark_Bay", "Exmouth_Gulf", "Pilbara", "Broome", "North_Kimberley")

pops <- fs::dir_ls(here("data", "popmaps"), glob = "*.txt") |>
    read_delim(
        delim = " ", 
        col_names = c("sample", "population"), 
        col_types = cols(), 
        id = "species"
    ) |>
    mutate(
        population = str_replace_all(population, "_", " "),
        species = str_remove(basename(species), "-.*")
    )

popmap_lat_long <- fs::dir_ls(
    here("data", "popmaps", "popmaps-lat-long"), 
    glob = "*.csv"
) |>
    purrr::map(read_csv, col_types = cols(), col_select = 1:7) |>
    bind_rows() |>
    unite(col = "sample", sep = "-", species, id, targetid) |>
    mutate(sample = str_replace_all(sample, " ", "-"))

ignore <- fs::dir_ls(here("results", "population-structure"), glob = "*.txt") |>
    read_tsv(
        col_names = "population",
        col_types = cols()
    ) |>
    distinct() |>
    pull() |>
    str_replace_all("_", " ")

# Problem samples to remove 
samples_remove <- read_tsv(
    here("data", "popmaps", "dropped-samples.tsv"), 
    skip = 1, 
    col_names = c("sample", "pop"),
    col_types = cols()
) |>
    pull(sample)
```

## Helper functions

This helper function is for generating the PCAs. It handles finding `x` and `y`
maxima/minima and formatting.

```{r helper-function-plotting}
# Function that takes the PCA list we'll generate below and plot the desired PCs
# for the specified population
plotPCA <- function(pca_dataframe, variance_dataframe, spec, WA_only, pcA, pcB, xstep = 1, ystep = 1, pal) {
    full_species <- set_names(c("ALA", "HMA", "HST"), c("Aipysurus laevis", "Hydrophis major", "Hydrophis stokesii"))
    plot_title <- names(full_species[full_species == spec])
    
    sp_pca <- pca_dataframe |>
        filter(species == spec, WA == WA_only)
    
    sp_var <- variance_dataframe |>
        filter(species == spec, WA == WA_only)
    
    # TODO: Figure out why this is being a piece of shit
    if(WA_only){
        wa_levels <- pca_dataframe |> 
            filter(WA == TRUE) |> 
            pull(population) |> 
            unique() |>
            as.character()
        
        sp_pca <- sp_pca |>
            mutate(
                population = factor(
                    population,
                    levels = wa_levels
                )
            )
    }
    
    # x-axis: max/min
    xmin <- sp_pca |>
        pull(pcA) |>
        min(na.rm = TRUE) |>
        floor()
    
    xmax <- sp_pca |>
        pull(pcA) |>
        max(na.rm = TRUE) |>
        ceiling()
    
    # y-axis: max/min
    ymin <- sp_pca |>
        pull(pcB) |>
        min(na.rm = TRUE) |>
        floor()
    
    ymax <- sp_pca |>
        pull(pcB) |>
        max(na.rm = TRUE) |>
        ceiling()
    
    # Variance strings
    pcA_var <- round(
        sp_var |>
            filter(pc_axes == pcA) |>
            pull(variance) * 100, digits = 2
    )
    
    pcB_var <- round(
        sp_var |>
            filter(pc_axes == pcB) |>
            pull(variance) * 100, digits = 2
    )
    
    sp_pca |>
        ggplot(
            aes(
                x = .data[[pcA]], y = .data[[pcB]],
                colour = population
            )
        ) +
        geom_point(size = 2.5, show.legend=TRUE) +
        scale_colour_manual(
            values = pal,
            drop = FALSE
        ) +
        labs(
            title = glue::glue("{plot_title}"),
            colour = "Population",
            x = glue::glue("{pcA} ({pcA_var}%)"),
            y = glue::glue("{pcB} ({pcB_var}%)")
        ) +
        scale_y_continuous(
            limits = c(ymin, ymax),
            breaks = seq(ymin, ymax, by = ystep),
            labels = as.character(seq(ymin, ymax, by = ystep)),
            expand = c(0, 0)
        ) +
        scale_x_continuous(
            limits = c(xmin, xmax),
            breaks = seq(xmin, xmax, by = xstep),
            labels = as.character(seq(xmin, xmax, by = xstep)),
            expand = c(0, 0)
        ) +
        theme(
            axis.text = element_text(size = 14),
            axis.title = element_text(size = 14),
            legend.position = "bottom",
            legend.title = element_text(hjust = 0.5, size = 14),
            legend.title.position = "top",
            legend.text = element_text(size = 14),
            plot.title = element_text(hjust = 0.5, face = "bold.italic")
        )
}
```

# Removed samples

A range of samples were excluded from the PCAs below. Further numerous populations
were ignored when running `STRUCTURE`. These can be found in the following files

- [dropped-samples.tsv][dropped]
- Per-species populations to ignore ([here][ignore])

## PCA

Principal component analysis (PCA) was conducted using the [Ipyrad analysis toolkit][ipatools].
The code responsible for the CSV files we're loading in below can be found
[here][pcaipynb]. Below, we're simply creating custom figures. We generate PCAs
for **all** samples across all populations for each species.

```{r load-pca-data}
pop_levels <- c(
    "Shark Bay" ,"Exmouth Gulf", "Pilbara", 
    "Broome", "North Kimberley", "Scott Reef", "Ashmore", "Heywood Shoal",
    "Gulf of Carpentaria", "North QLD", "South QLD", "New Caledonia"
)

# PCA dataframes
df_pca <- fs::dir_ls(here("results", "population-structure", "pca"), glob = "*.csv") |>
    (\(x) x[!str_detect(x, "variance")])() |>
    imap(\(x, i) {
        fname <- str_remove(basename(i), ".csv")
        spec <- str_remove(fname, "-.*")
        wa <- if(str_detect(fname, "WA")) TRUE else FALSE
        
        tmp <- x |>
            read_csv(col_names = TRUE, col_types = cols())
        colnames(tmp) <- c("sample", paste0("PC", 1:(ncol(tmp) - 1)))
        pops |>
            filter(species == spec) |>
            inner_join(tmp) |>
            mutate(WA = wa, .after = species)
    }) |>
    list_rbind() |>
    select(species, WA, sample, population, PC1, PC2, PC3) |>
    mutate(
        population = str_replace_all(population, "_", " "),
        population = factor(population, levels = pop_levels)
    )

# Variance dataframes
df_variance <- fs::dir_ls(here("results", "population-structure", "pca"), glob = "*variance.csv") |>
    imap(\(x, i) {
        fname <- str_remove(basename(i), ".csv")
        spec <- str_remove(fname, "-.*")
        wa <- if(str_detect(fname, "WA")) TRUE else FALSE
        x |>
            read_csv(
                skip = 1,
                col_names = c("pc_axes", "variance"),
                col_types = cols()
            ) |>
            mutate(
                species = spec,
                WA = wa,
                pc_axes = pc_axes + 1,
                pc_axes = paste0("PC", pc_axes)
            ) |>
            filter(pc_axes %in% c("PC1", "PC2", "PC3"))
    }) |>
    list_rbind()
```

```{r get-legend}
# Custom ordering of colour palette
pal_PCA <- c(
    "#DCB0F2FF","#F89C74FF", "#F6CF71FF", "#66C5CCFF","#FE88B1FF", "#87C55FFF",
    "#B497E7FF", "#B3B3B3FF", "#C9DB74FF", "#9EB9F3FF", "#D3B484FF", "#8BE0A4FF"
)
names(pal_PCA) <- pop_levels
```

### All samples

```{r pca-ALA}
ala_pc_1_2 <- plotPCA(pca_dataframe = df_pca, variance_dataframe = df_variance, WA_only = FALSE, spec = "ALA", pcA = "PC1", pcB = "PC2", xstep = 2, pal = pal_PCA)
hst_pc_1_2 <- plotPCA(pca_dataframe = df_pca, variance_dataframe = df_variance, WA_only = FALSE, spec = "HMA", pcA = "PC1", pcB = "PC2", xstep = 2, ystep = 2, pal = pal_PCA)
hma_pc_1_2 <- plotPCA(pca_dataframe = df_pca, variance_dataframe = df_variance, WA_only = FALSE, spec = "HST", pcA = "PC1", pcB = "PC2", xstep = 2, ystep = 2, pal = pal_PCA)

design <- "12\n34"

plot_pca_all <- ala_pc_1_2 + hst_pc_1_2 + hma_pc_1_2 + guide_area() +
    plot_layout(
        design = design,
        guides = "collect",
        axes = "collect"
    ) +
    plot_annotation(tag_levels = 'A') &
    guides(colour = guide_legend(ncol = 2, override.aes = list(size = 3))) &
    theme(plot.tag = element_text(size = 14, face = "bold"))

ragg::agg_png(
    filename = here("results", "population-structure", "pca-all-species-all-pops.png"),
    width = 1500, 
    height = 1500,
    units = "px",  
    res = 150
)
plot_pca_all
invisible(dev.off())

plot_pca_all
```

### WA Coast

And the same for *just* the WA coastline samples

```{r}
ala_pc_1_2 <- plotPCA(pca_dataframe = df_pca, variance_dataframe = df_variance, WA_only = TRUE, spec = "ALA", pcA = "PC1", pcB = "PC2", xstep = 2, ystep = 2, pal = pal_PCA)
hma_pc_1_2 <- plotPCA(pca_dataframe = df_pca, variance_dataframe = df_variance, WA_only = TRUE, spec = "HMA", pcA = "PC1", pcB = "PC2", xstep = 2, ystep = 2, pal = pal_PCA)
hst_pc_1_2 <- plotPCA(pca_dataframe = df_pca, variance_dataframe = df_variance, WA_only = TRUE, spec = "HST", pcA = "PC1", pcB = "PC2", xstep = 2, ystep = 2, pal = pal_PCA)

design <- "12\n34"

plot_pca_wa <- ala_pc_1_2 + hma_pc_1_2 + hst_pc_1_2 + guide_area() +
    plot_layout(
        design = design,
        guides = "collect",
        axes = "collect"
    ) &
    plot_annotation(tag_levels = 'A') &
    guides(colour = guide_legend(ncol = 2, override.aes = list(size = 3))) &
    theme(plot.tag = element_text(size = 14, face = "bold"))

ragg::agg_png(
    filename = here("results", "population-structure", "pca-all-species-WA-pops.png"),
    width = 1500, 
    height = 1500,
    units = "px",  
    res = 150
)
plot_pca_wa
invisible(dev.off())

plot_pca_wa
```

We can see that the Shark Bay samples partition off in both *A. laevis* and
*H. major*.

## STRUCTURE

`STRUCTURE` plots are then generated for WA populations, specifically: Broome,
Pilbara, Exmouth Gulf, Shark Bay and North Kimberley. In the PCAs above, samples
from these populations typically cluster together, showing little structure.
Consequently, we're using this method as a more granular approach to determine
if there is any allele sharing between the populations.

```{r read-structure}
structure_df <- fs::dir_ls(
    here("results", "population-structure", "structure-all-updated"), 
    glob = "*plot.csv", 
    recurse = TRUE
) |>
    map(\(x) {
        tmp <- x |>
            read_csv(col_names = TRUE, col_types = cols())
        tmp |>
            mutate(K = glue::glue("K{ncol(tmp) - 1}")) |>
            pivot_longer(names_to = "partition", values_to = "proportion", 2:ncol(tmp)) |>
            mutate(partition = as.numeric(partition) + 1) |>
            rename(sample = `...1`) |>
            left_join(pops) |>
            left_join(popmap_lat_long)
    }) |>
    bind_rows() |>
    mutate(
        partition = forcats::fct_inorder(as.character(partition)),
        population = str_replace_all(population, "_", " "),
        # population = factor(population, levels = structure_pops),
        .by = K
    ) |>
    arrange(species, longitude) |>
    mutate(
        sample = forcats::fct_inorder(sample),
        population = forcats::fct_inorder(population)
    )

# Custom ordering of colour palette
# pal <- c(
#     "#ff595e", "#ffca3a", "#8ac926", "#1982c4"
# )
pal <- c("#1982c4", "#1982c480", "#1982c440", "#1982c410")
names(pal) <- 1:4
```

### Choosing the correct K (N-ancestral populations)

Choosing the best `K` value can be determined by plotting the estimated
log-probability for each `K` value, in addition to the delta-K value. The values
can be interpreted as follows:

- **Estimated log-probability**: Higher values equal better model fit.
- **Delta-K**: The rate of change between successive K-values. Larger values
indicate greater change in the model fit.

The results for the three species are shown below.

```{r}
etable <- fs::dir_ls(
    here("results", "population-structure", "structure-all-updated"), 
    glob = "*etable.csv",
    recurse = TRUE
) |>
    read_csv(col_names = TRUE, col_types = cols(), id = "species") |>
    mutate(species = str_remove(basename(species), "-.*")) |>
    rename(K = `...1`)

etable |>
    gt(groupname_col = "species") |>
    fmt_number(
        columns = c("K", "Nreps"),
        drop_trailing_zeros = TRUE
    ) |>
    fmt_number(
        columns = 3:7,
        n_sigfig = 4
    ) |>
    as_raw_html()
```

Plotting the table above looks like the following.

```{r}
prb <- etable |>
    ggplot(aes(x = K, y = estLnProbMean, colour = species)) +
    geom_point() +
    geom_line() +
    scale_y_continuous(
        name = "Estimated mean log-probability",
        breaks = seq(0, -105000, -10000),
        limits = c(-105e3, -2e4),
        labels = scales::label_number(style_negative = "minus")
    ) +
    scale_x_continuous(breaks = seq(2, 9, 1))

dk <- etable |>
    ggplot(aes(x = K, y = deltaK, colour = species)) +
    geom_point() +
    geom_line() +
    scale_x_continuous(breaks = seq(2, 9, 1)) +
    scale_y_continuous(
        name = "Delta-K",
        breaks = seq(0, 15, 1)
    )

plot_models <- prb/dk +
    plot_layout(axes = "collect", guides = "collect")

ragg::agg_png(
    filename = here("results", "population-structure", "structure-model-fit.png"),
    width = 1000, 
    height = 800,
    units = "px",  
    res = 150
)
plot_models
invisible(dev.off())

plot_models
```

From the figures above, it looks like `K = 4` is the best model fit in *Aipysurus*
*laevis* and *Hydrophis stokesii*, while `K = 3` appears the be the best fit for
*Hydrophis major*.

### *Aipysurus laevis*

```{r}
pop_groups <- structure_df |>
    filter(species == "ALA", K == "K4") |> 
    select(sample, population, species) |>
    distinct() |>
    mutate(size = 1) |>
    ggplot(aes(x = sample, y = size, fill = population)) +
    geom_col(width = 1) +
    scale_fill_manual(values = pal_PCA) +
    scale_y_continuous(expand = c(0,0)) +
    labs(x = NULL) + 
    theme(
        axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.98),
    ) +
    theme(
        legend.position = "top",
        legend.title = element_blank()
    )

plot_structure_ALA <- structure_df |>
    filter(species == "ALA", K =="K4") |>
    ggplot(aes(x = sample, y = proportion, fill = partition)) +
    geom_col(position = "stack") +
    scale_fill_manual(values = pal) +
    scale_y_continuous(
        limits = c(0, 1),
        breaks = c(0, 0.5, 1),
        labels = as.character(seq(0, 1, 0.5)),
        expand = c(0, 0)
    ) +
    labs(x = NULL) +
    guides(fill = guide_legend(title.position = "bottom")) +
    theme(
        axis.title.y = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.98),
        legend.position = "bottom",
        legend.title = element_text(hjust = 0.5),
        strip.text.x = element_text(
            face = "bold", 
            hjust = 0.5,
            size = 14,
            angle = 90
        ),
        strip.text.y = element_text(size = 14, hjust = 0.5),
        strip.background = element_rect(fill='transparent', colour = NA),
        panel.spacing.x = unit(0.4, "line"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()
    )

plot_structure_ALA <- pop_groups / plot_structure_ALA +
    plot_layout(heights = c(1,4), axes = "collect") &
    theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()
    )

ragg::agg_png(
    filename = here("results", "population-structure", "structure-ALA.png"),
    width = 1500, 
    height = 1000,
    units = "px",  
    res = 150
)
plot_structure_ALA
invisible(dev.off())

plot_structure_ALA
```

### *Hydrophis major*

```{r}
pop_groups <- structure_df |>
    filter(species == "HMA", K == "K3") |> 
    select(sample, population, species) |>
    distinct() |>
    mutate(size = 1) |>
    ggplot(aes(x = sample, y = size, fill = population)) +
    geom_col(width = 1) +
    scale_fill_manual(values = pal_PCA) +
    scale_y_continuous(expand = c(0,0)) +
    labs(x = NULL) + 
    theme(
        axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.98),
    ) +
    theme(
        legend.position = "top",
        legend.title = element_blank()
    )

plot_structure_HMA <- structure_df |>
    filter(species == "HMA", K == "K3") |>
    ggplot(aes(x = sample, y = proportion, fill = partition)) +
    geom_col(position = "stack") +
    scale_fill_manual(values = pal) +
    scale_y_continuous(
        breaks = c(0, 0.5, 1),
        labels = as.character(seq(0, 1, 0.5)),
        expand = c(0, 0)
    ) +
    labs(x = NULL) + 
    guides(fill = guide_legend(title.position = "bottom")) +
    theme(
        axis.title.y = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.98),
        legend.position = "bottom",
        legend.title = element_text(hjust = 0.5),
        strip.text.x = element_text(
            face = "bold", 
            hjust = 0.5,
            size = 14,
            angle = 45
        ),
        strip.text.y = element_text(size = 14, hjust = 0.5),
        strip.background = element_rect(fill='transparent', colour = NA),
        panel.spacing.x = unit(0.4, "line"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()
    )

plot_structure_HMA <- pop_groups / plot_structure_HMA +
    plot_layout(heights = c(1,4), axes = "collect") &
    theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()
    )

ragg::agg_png(
    filename = here("results", "population-structure", "structure-HMA.png"),
    width = 1500, 
    height = 1000,
    units = "px",  
    res = 150
)
plot_structure_HMA
invisible(dev.off())

plot_structure_HMA
```

### *Hydrophis stokesii*

```{r}
pop_groups <- structure_df |>
    filter(species == "HST", K == "K4") |> 
    select(sample, population, species) |>
    distinct() |>
    mutate(size = 1) |>
    ggplot(aes(x = sample, y = size, fill = population)) +
    geom_col(width = 1) +
    scale_fill_manual(values = pal_PCA) +
    scale_y_continuous(expand = c(0,0)) +
    labs(x = NULL) + 
    theme(
        axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.98),
    ) +
    theme(
        legend.position = "top",
        legend.title = element_blank()
    )

plot_structure_HST <- structure_df |>
    filter(species == "HST", K == "K4") |>
    ggplot(aes(x = sample, y = proportion, fill = partition)) +
    geom_col(position = "stack") +
    scale_fill_manual(values = pal) +
    scale_y_continuous(
        breaks = c(0, 0.5, 1),
        labels = as.character(seq(0, 1, 0.5)),
        expand = c(0, 0)
    ) +
    labs(x = NULL) + 
    guides(fill = guide_legend(title.position = "bottom")) +
    theme(
        axis.title.y = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.98),
        legend.position = "bottom",
        legend.title = element_text(hjust = 0.5),
        strip.text.x = element_text(
            face = "bold", 
            hjust = 0.1,
            vjust = 0.5,
            size = 14,
            angle = 90
        ),
        strip.text.y = element_text(size = 14, hjust = 0.5),
        strip.background = element_rect(fill='transparent', colour = NA),
        panel.spacing.x = unit(0.4, "line"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()
    )

plot_structure_HST <- pop_groups / plot_structure_HST +
    plot_layout(heights = c(1,4), axes = "collect") &
    theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()
    )

ragg::agg_png(
    filename = here("results", "population-structure", "structure-HST.png"),
    width = 1500, 
    height = 1000,
    units = "px",  
    res = 150
)
plot_structure_HST
invisible(dev.off())

plot_structure_HST
```

## STRUCTURE: WA coast samples

```{r}
structure_df_wa <- fs::dir_ls(
    here("results", "population-structure", "structure"), 
    glob = "*plot.csv", 
    recurse = TRUE
) |>
    map(\(x) {
        tmp <- x |>
            read_csv(col_names = TRUE, col_types = cols())
        tmp |>
            mutate(K = glue::glue("K{ncol(tmp) - 1}")) |>
            pivot_longer(names_to = "partition", values_to = "proportion", 2:ncol(tmp)) |>
            mutate(partition = as.numeric(partition) + 1) |>
            rename(sample = `...1`) |>
            left_join(pops) |>
            left_join(popmap_lat_long)
    }) |>
    bind_rows() |>
    mutate(
        partition = forcats::fct_inorder(as.character(partition)),
        population = str_replace_all(population, "_", " "),
        population = factor(pop, levels = wa_pops),
        .by = K
    ) |>
    arrange(species, longitude) |>
    mutate(sample = forcats::fct_inorder(sample))
```

### *Aipysurus laevis*

```{r}
pop_groups <- structure_df_wa |>
    filter(species == "ALA", K == "K3") |> 
    select(sample, population = pop, species) |>
    distinct() |>
    mutate(
        size = 1,
        population = str_replace_all(population, "_", " ")
    ) |>
    ggplot(aes(x = sample, y = size, fill = population)) +
    geom_col(width = 1) +
    scale_fill_manual(values = pal_PCA) +
    scale_y_continuous(expand = c(0,0)) +
    labs(x = NULL) + 
    theme(
        axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.98),
    ) +
    theme(
        legend.position = "top",
        legend.title = element_blank()
    )

plot_structure_ALA_wa <- structure_df_wa |>
    filter(species == "ALA", K == "K3") |>
    ggplot(aes(x = sample, y = proportion, fill = partition)) +
    geom_col(position = "stack") +
    scale_fill_manual(values = pal) +
    scale_y_continuous(
        breaks = c(0, 0.5, 1),
        labels = as.character(seq(0, 1, 0.5)),
        expand = c(0, 0)
    ) +
    labs(x = NULL) + 
    guides(fill = guide_legend(title.position = "bottom")) +
    theme(
        axis.title.y = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.98),
        legend.position = "bottom",
        legend.title = element_text(hjust = 0.5),
        strip.text.x = element_text(
            face = "bold", 
            hjust = 0.1,
            vjust = 0.5,
            size = 14,
            angle = 90
        ),
        strip.text.y = element_text(size = 14, hjust = 0.5),
        strip.background = element_rect(fill='transparent', colour = NA),
        panel.spacing.x = unit(0.4, "line"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()
    )

plot_structure_ALA_wa <- pop_groups / plot_structure_ALA_wa +
    plot_layout(heights = c(1,4), axes = "collect") &
    theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()
    )

ragg::agg_png(
    filename = here("results", "population-structure", "structure-ALA-WA.png"),
    width = 1500, 
    height = 1000,
    units = "px",  
    res = 150
)
plot_structure_ALA_wa
invisible(dev.off())

plot_structure_ALA_wa
```

### *Hydrophis major*

```{r}
pop_groups <- structure_df_wa |>
    filter(species == "HMA", K == "K3") |> 
    select(sample, population = pop, species) |>
    distinct() |>
    mutate(
        size = 1,
        population = str_replace_all(population, "_", " ")
    ) |>
    ggplot(aes(x = sample, y = size, fill = population)) +
    geom_col(width = 1) +
    scale_fill_manual(values = pal_PCA) +
    scale_y_continuous(expand = c(0,0)) +
    labs(x = NULL) + 
    theme(
        axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.98),
    ) +
    theme(
        legend.position = "top",
        legend.title = element_blank()
    )

plot_structure_HMA_wa <- structure_df_wa |>
    filter(species == "HMA", K == "K3") |>
    ggplot(aes(x = sample, y = proportion, fill = partition)) +
    geom_col(position = "stack") +
    scale_fill_manual(values = pal) +
    scale_y_continuous(
        breaks = c(0, 0.5, 1),
        labels = as.character(seq(0, 1, 0.5)),
        expand = c(0, 0)
    ) +
    labs(x = NULL) + 
    guides(fill = guide_legend(title.position = "bottom")) +
    theme(
        axis.title.y = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.98),
        legend.position = "bottom",
        legend.title = element_text(hjust = 0.5),
        strip.text.x = element_text(
            face = "bold", 
            hjust = 0.1,
            vjust = 0.5,
            size = 14,
            angle = 90
        ),
        strip.text.y = element_text(size = 14, hjust = 0.5),
        strip.background = element_rect(fill='transparent', colour = NA),
        panel.spacing.x = unit(0.4, "line"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()
    )

plot_structure_HMA_wa <- pop_groups / plot_structure_HMA_wa +
    plot_layout(heights = c(1,4), axes = "collect") &
    theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()
    )

ragg::agg_png(
    filename = here("results", "population-structure", "structure-HMA-WA.png"),
    width = 1500, 
    height = 1000,
    units = "px",  
    res = 150
)
plot_structure_HMA_wa
invisible(dev.off())

plot_structure_HMA_wa
```

### *Hydrophis stokesii*

```{r}
pop_groups <- structure_df_wa |>
    filter(species == "HST", K == "K3") |> 
    select(sample, population = pop, species) |>
    distinct() |>
    mutate(
        size = 1,
        population = str_replace_all(population, "_", " ")
    ) |>
    ggplot(aes(x = sample, y = size, fill = population)) +
    geom_col(width = 1) +
    scale_fill_manual(values = pal_PCA) +
    scale_y_continuous(expand = c(0,0)) +
    labs(x = NULL) + 
    theme(
        axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.98),
    ) +
    theme(
        legend.position = "top",
        legend.title = element_blank()
    )

plot_structure_HST_wa <- structure_df_wa |>
    filter(species == "HST", K == "K3") |>
    ggplot(aes(x = sample, y = proportion, fill = partition)) +
    geom_col(position = "stack") +
    scale_fill_manual(values = pal) +
    scale_y_continuous(
        breaks = c(0, 0.5, 1),
        labels = as.character(seq(0, 1, 0.5)),
        expand = c(0, 0)
    ) +
    labs(x = NULL) + 
    guides(fill = guide_legend(title.position = "bottom")) +
    theme(
        axis.title.y = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.98),
        legend.position = "bottom",
        legend.title = element_text(hjust = 0.5),
        strip.text.x = element_text(
            face = "bold", 
            hjust = 0.1,
            vjust = 0.5,
            size = 14,
            angle = 90
        ),
        strip.text.y = element_text(size = 14, hjust = 0.5),
        strip.background = element_rect(fill='transparent', colour = NA),
        panel.spacing.x = unit(0.4, "line"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()
    )

plot_structure_HST_wa <- pop_groups / plot_structure_HST_wa +
    plot_layout(heights = c(1,4), axes = "collect") &
    theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()
    )

ragg::agg_png(
    filename = here("results", "population-structure", "structure-HST-WA.png"),
    width = 1500, 
    height = 1000,
    units = "px",  
    res = 150
)
plot_structure_HST_wa
invisible(dev.off())

plot_structure_HST_wa
```

# Isolation by Distance (IBD)

Lastly, we'll perform Isolation-by-distance analysis. We'll first load the VCF
data and estimate the genetic distance matrix (Nei's genetic distance).

Note: I also found this repo helpful - https://github.com/jdalapicolla/IBD_models.R

```{r}
# Load the snpR library and the custom scripts (found in the scripts directory)
library(snpR)
library(marmap)
library(fossil)
library(vegan)
source(here("scripts", "snpR-edited.R"))

# Load updated lat/long popmaps
set.seed(123)
meta_latLong <- fs::dir_ls(here("data", "IBD-popmaps"), glob = "*avg.csv") |>
    (\(x) set_names(x, str_remove(basename(x), "-.*")))() |> 
    purrr::map(\(x) {
        x |>
            read_csv(col_types = cols()) |>
            select(species, id, targetid, pop, latitude, longitude) |>
            unite(col = "sampID", sep = "-", species, id, targetid) |>
            filter(!is.na(latitude), pop %in% wa_pops, ! sampID %in% samples_remove) |>
            select(sampID, pop, x = longitude, y = latitude) |>
            # For snakes with identical long/lat + population, choose one at random
            slice_sample(n = 1, by = c(pop, x, y))
    })

# Load the VCF files as snpR data objects
vcfs <- fs::dir_ls(
    here("results", "ipyrad"),
    glob = "*stringent.vcf.gz",
    recurse = TRUE
) |>
    (\(x) set_names(x, str_remove(basename(x), "-stringent.*")))() |>
    imap(\(v,i) {
        filter_val <- case_when(
            i == "ALA" ~ 0.9, 
            i == "HMA" ~ 0.9, 
            i == "HST" ~ 0.95
        )
        tmp_vcf <- read_vcf(file = v)
        
        filt_vcf <- filter_snps(
            tmp_vcf,
            min_ind = filter_val,
            re_run = "full"
        )
        
        # Subset for samples to keep + add in metadata
        filt_vcf <- subset_snpR_data(
            x = filt_vcf, 
            .facets = "sampID",
            .subfacets = meta_latLong[[i]]$sampID
        )
        
        sample.meta(filt_vcf) <- left_join(sample.meta(filt_vcf), meta_latLong[[i]])
        
        filt_vcf <- custom_calc_genetic_distances(
            filt_vcf, 
            facets = c("sampID", "pop"), 
            method = "Nei"
        )
        filt_vcf
    })
```

## Mantel statistics and spatial auto-correlation

Next we'll perform the Mantel tests and generate the autocorrelation plots. We're
using `vegan` to perform Mantel tests as this package has the autocorrelation
function as well. Consequently, we're not using the in-built IBD function in
`snpR`.

```{r mantel-tests}
mantel_results <- vcfs |>
    imap(\(x, i) {
        genMat <- get.snpR.stats(x, facets = "sampID", stats = "genetic_distance")
        genMat <-as.matrix(genMat$sampID$.base$Nei)
        
        lat_long <- meta_latLong[[i]] |>
            select(3, 4)
        
        if(i == "ALA") {
            WAcoast <- getNOAA.bathy(
                lon1 = 111.7, lon2 = 127.2,
                lat1 = -26.8, lat2 = -9.9,
                resolution = 3, 
                keep = TRUE
            )
        } else if(i == "HMA") {
            WAcoast <- getNOAA.bathy(
                lon1 = 111.7, lon2 = 127.2,
                lat1 = -26.8, lat2 = -9.9, 
                resolution = 3, keep = TRUE
            )
        } else {
            WAcoast <- getNOAA.bathy(
                lon1 = 111.7, lon2 = 127.2,
                lat1 = -26.8, lat2 = -9.9, 
                resolution = 3, 
                keep = TRUE
            )
        }
        
        transMat <- trans.mat(WAcoast) 
        lcDist <- lc.dist(transMat, lat_long, res = "dist")
        lcDist <- as.matrix(lcDist)
        
        rownames(lcDist) <- meta_latLong[[i]]$sampID
        colnames(lcDist) <- meta_latLong[[i]]$sampID
        
        # Arrange geographic distance matrix by genetic distance matrix
        lcDist <- lcDist[rownames(genMat), colnames(genMat)]
        
        # Grouping data into 200km bins - increases amount of data for correlation
        classes <- seq(0, round( (max(lcDist, na.rm = TRUE) + 50), digits = -2 ), 200)
        
        # Example of running the Mantel statistic using vegan
        output.mantel <- vegan::mantel(
            as.dist(genMat),
            as.dist(lcDist),
            permutations = 10000
        )
        
        output.mantel.summary <- tibble(
            "Observed correlation" = output.mantel$statistic,
            "Significance" = output.mantel$signif,
            "Replicates" = output.mantel$permutations
        )
        
        # Correlog analysis - binned data
        output.mantel.correlog <- vegan::mantel.correlog(
            D.eco = as.dist(genMat),
            D.geo = as.dist(lcDist),
            break.pts = classes,
            nperm = 10000,
            progressive = TRUE,
            cutoff = FALSE
        )
        
        list(
            "mantel" = output.mantel.summary,
            "correlog" = output.mantel.correlog,
            "genetic_dist" = genMat,
            "geo_dist" = lcDist
        )
    })

mantel_results |>
    purrr::map(\(x) {
        x$mantel
    }) |>
    bind_rows(.id = "Species") |>
    gt() |>
    fmt_number(
        columns = c("Observed correlation", "Significance"),
        decimals = 4
    ) |>
    fmt_number(
        columns = "Replicates", use_seps = TRUE, drop_trailing_zeros = TRUE
    ) |>
    as_raw_html()
```

## IBD visualisation

### Geographic distance vs Genetic distance

Lastly, let's generate a scatter plot of distance vs genetic distance. Add the
correlation coefficients and p-values in Inkscape or something.

```{r}
df_gen_geo <- mantel_results |>
    imap(\(x, i) {
        genDist <- x[["genetic_dist"]]
        geoDist <- x[["geo_dist"]]
        
        # Reformat genetic distance matrix
        genDist[upper.tri(genDist, diag = T)] = NA
        genDist <- genDist |>
            reshape2::melt() |>
            na.omit() |>
            arrange(Var1) |>
            setNames(c("X1", "X2","GenDist")) |>
            as_tibble()
        
        # Reformat geographic distance matrix
        geoDist[upper.tri(geoDist, diag = T)] = NA
        geoDist <- geoDist |>
            reshape2::melt() |>
            na.omit() |>
            arrange(Var1) |>
            setNames(c("X1", "X2","GeoDist")) |>
            as_tibble()
        
        left_join(genDist, geoDist) |>
            na.omit()
    }) |>
    bind_rows(.id = "species") |>
    mutate(species = case_when(
        species == "ALA" ~ "Aipysurus laevis",
        species == "HMA" ~ "Hydrophis major",
        species == "HST" ~ "Hydrophis stokesii"
    ))

# Plot distance (kilometres) vs genetic distance (Nei)
plot_scatter <- df_gen_geo |>
    mutate(species = factor(species, levels = c("Aipysurus laevis", "Hydrophis major", "Hydrophis stokesii"))) |>
    ggplot(aes(x = GeoDist, y = GenDist, colour = species)) +
    geom_point(alpha = 0.4) +
    geom_smooth(mapping = aes(colour = species), method = "lm", show.legend = FALSE) +
    scale_x_continuous(
        name = "Distance (Km)",
        breaks = seq(0, 1600, 200),
        labels = as.character(seq(0, 1600, 200)),
        expand = c(0.01, 0, 0.02, 0)
    ) +
    scale_y_continuous(
        name = "Genetic distance (Nei)",
        breaks = seq(0, 0.055, 0.005),
        labels = as.character(seq(0, 0.055, 0.005)),
        limits = c(0, 0.055),
        expand = c(0.02, 0)
    ) +
    scale_colour_brewer(palette = "Set1") +
    guides(colour = guide_legend(override.aes = list(size = 6))) +
    theme(
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        legend.position = "top",
        legend.text = element_text(size = 14, face = "italic"),
        legend.title = element_blank()
    )

cairo_pdf(
    here("results", "population-structure", "mantel-scatter.pdf"),
    width = 12, height = 12
)
plot_scatter
invisible(dev.off())

plot_scatter
``` 

### Spatial auto-correlation

```{r}
plots <- mantel_results |>
    purrr::imap(\(x, i) {
        spc <- case_when(
            i == "ALA" ~ "Aipysurus laevis",
            i == "HMA" ~ "Hydrophis major",
            i == "HST" ~ "Hydrophis stokesii"
        )
        tmp <- x$correlog$mantel.res |>
            as_tibble()
        
        tmp |>
            mutate(
                classification = case_when(
                    `Mantel.cor` < 0 & `Pr(corrected)` < 0.05 ~ "Lower",
                    `Mantel.cor` > 0 & `Pr(corrected)` < 0.05 ~ "Higher",
                    .default = "Neutral"
                ),
                classification = factor(classification, levels = c("Higher", "Neutral", "Lower"))
            ) |>
            ggplot(aes(x = class.index, y = Mantel.cor)) +
            geom_hline(yintercept = 0, colour = "red") +
            geom_line(colour = "black") +
            geom_point(
                mapping = aes(fill = classification), 
                shape = 22, 
                size = 4, 
                show.legend = TRUE, 
                colour = "black"
            ) +
            labs(
                x = "Least-cost distance (KM)",
                y = "Mantel correlation",
                fill = NULL,
                title = spc
            ) +
            scale_fill_manual(values = c("#00BA38", "grey80", "#619CFF"), drop = FALSE) +
            scale_y_continuous(
                limits = c(-0.3, 0.2),
                breaks = seq(-0.3, 0.2, 0.1),
                labels = c("0.3", "-0.2", "-0.1", "0", "0.1", "0.2"),
                expand = c(0.01, 0)
            ) +
            scale_x_continuous(
                limits = c(0, (max(tmp$class.index) + 100) ),
                breaks = seq(0, ( max(tmp$class.index) + 100), 200 ),
                labels = seq(0, ( max(tmp$class.index) + 100), 200 ),
                expand = c(0.01, 0)
            ) +
            guides(fill = guide_legend(override.aes = list(size = 6))) +
            theme(
                axis.text = element_text(size = 14),
                axis.title = element_text(size = 16),
                legend.text = element_text(size = 14),
                plot.title = element_text(face = "italic", hjust = 0.5)
            )
    })

design <- "12\n34"

plot_correlog <- plots$ALA + plots$HMA + plots$HST + guide_area() +
    plot_layout(
        design = design,
        guides = "collect",
        axes = "collect"
    )

cairo_pdf(
    here("results", "population-structure", "mantel-correlog.pdf"),
    width = 12, height = 12
)
plot_correlog
invisible(dev.off())

plot_correlog
```


