---
title: "Diversity statistics"
date: "`r Sys.Date()`"
output: 
    github_document:
        toc: true
---

This directory contains results relating to diversity statistics, both within
and between populations. The statistics are generated from the unfiltered
`Ipyrad` VCFs, with all filtering being applied by the package `snpR`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE, message = FALSE,
    warning = FALSE, error = FALSE,
    out.width = "100%", fig.width = 12, fig.height = 12
)

suppressPackageStartupMessages({
    library(snpR)
    library(tidyverse)
    library(here)
    library(gt)
    library(patchwork)
})

# Source custom genetic distance function
source(here("scripts", "snpR-edited.R"))

# Problem samples to remove 
samples_remove <- read_tsv(
    here("data", "popmaps", "dropped-samples.tsv"), 
    skip = 1, 
    col_names = c("sample", "pop"),
    col_types = cols()
) |>
    pull(sample)

theme_set(theme_bw())

fs::dir_create(path = here("results", "diversity-statistics"))

pop_levels <- c(
    "Shark_Bay", "Coastal_WA", "Offshore_WA", 
    "Gulf_of_Carpentaria", "New_Caledonia"
)
```

## Load metadata and VCF files

Here we're simply loading the population-map files and storing them into the
object `meta`. This object is a named `list`, where each list element is a
dataframe.

Note that we're going to remove the North and South QLD samples, as there aren't
many individuals in these groups and we're not certain of the structure in these
Eastern populations. As such, we're just keeping the well sampled Gulf of
Carpentaria population where available.

```{r load-meta}
# Load data
meta <- fs::dir_ls(
    here("data", "popmaps"),
    glob = "*.txt"
) |>
    (\(x) set_names(x, str_remove(basename(x), "-.*")))() |>
    map(\(x) {
        x |>
            read_delim(
                delim = " ", 
                col_names = c("sampID", "population"), 
                col_types = cols()
            ) |>
            filter(
                ! str_detect(population, "QLD"), 
                ! sampID %in% samples_remove
            )
    })
```

### Adapting metadata

Based off the PCA analyses, we're going to regroup some samples. Specifically,
all WA coastline samples *except* Shark Bay are going to be clustered into
"Coastal WA". Additionally, offshore reef samples are grouped as "Offshore WA".

```{r regroup-samples}
meta <- meta |>
    imap(\(x, i) {
        if(i == "ALA") {
            x <- x |>
                mutate(
                    pop = case_when(
                        population %in% c("Broome", "Pilbara", "Exmouth_Gulf") ~ "Coastal_WA",
                        population %in% c("Ashmore", "Heywood_Shoal", "Scott_Reef") ~ "Offshore_WA",
                        .default = population
                    )
                )
        }
        
        if(i == "HMA") {
            x <- x |>
                mutate(
                    pop = case_when(
                        population %in% c("Broome", "Pilbara", "Exmouth_Gulf") ~ "Coastal_WA",
                        .default = population
                    )
                )
        }
        
        if(i == "HST") {
            x <- x |>
                mutate(
                    pop = case_when(
                        population %in% c("Broome", "Pilbara", "Exmouth_Gulf", "North_Kimberley") ~ "Coastal_WA",
                        population %in% c("Ashmore", "Heywood_Shoal") ~ "Offshore_WA",
                        .default = population
                    )
                )
        }
        
        return(x)
    })
```

### Load VCF files

Next we load the VCF files into `R` as `snpRdata` objects. Similarly, we load
each species' file separately and store it in a list. Filtering is applied here
so we don't have to pre-filter our data ahead of time.

We also apply some sample filtering (i.e. removing QLD samples from snpRobject)
and add in the metadata after loading the data. This is to ensure the correct
metadata is assigned to the right sample.

```{r load-vcfs, results='hide'}
# Get each species' VCF file
vcfs <- fs::dir_ls(
    here("results", "ipyrad"),
    glob = "*stringent.vcf.gz",
    recurse = TRUE
) |>
    (\(x) set_names(x, str_remove(basename(x), "-stringent.*")))() |>
    imap(\(v,i) {
        tmp_vcf <- read_vcf(file = v)
        
        filt_vcf <- filter_snps(
            tmp_vcf,
            min_ind = 0.75,
            re_run = "full"
        )
        
        # Subset for samples to keep + add in metadata
        filt_vcf <- subset_snpR_data(
            x = filt_vcf, 
            .facets = "sampID",
            .subfacets = meta[[i]]$sampID
        )
        
        sample.meta(filt_vcf) <- left_join(sample.meta(filt_vcf), meta[[i]]) |>
            mutate(pop = factor(pop, levels = pop_levels))
        filt_vcf
    })
```

## Calculate statistics

Next we'll calculate a range of diversity statistics. The program `snpR` has a
really convenient function `calc_basic_snp_stats()` which practically estimates
all the statistics we want.

Below we estimate π, H~o~, H~e~, F~IS~, HWE and pairwise F~ST~. We estimate
all statistics by population.

```{r calc-basic-stats}
## WA all
# Calculate within population statistics
stats_within <- vcfs |>
    map(\(v) {
        tmp <- calc_basic_snp_stats(v, facets = "pop", fst.method = "WC")
        get.snpR.stats(tmp, facets = "pop", stats = c("pi", "hwe", "ho", "he", "fis", "fst"))
    })

# Divergence statistics
stats_between <- vcfs |>
    map(\(v) {
        tmp <- custom_calc_genetic_distances(v, facets = "pop", method = "Nei")
        get.snpR.stats(tmp, facets = "pop", stats = "genetic_distance")
    })
```

### Calculate 

We used a separate program to calculate nucleotide diversity (π) in each population.
This program is `pixy`, which uses invariant data in its calculation in addition
to variant sites. The authors of `Ipyrad` wrote a script that can return an
**all-sites** VCF file from the `.loci` files. I've passed this VCF to `pixy` to
get a more accurate estimation of π in each population.

```{r}
pixy_pi <- fs::dir_ls(
    here("results"),
    glob = "*pi.txt",
    recurse = TRUE
) |>
    read_tsv(col_types = cols(), id = "species") |>
    rename(population = pop) |>
    filter(!str_detect(population, "QLD")) |>
    mutate(
        species = str_remove(basename(species), "_.*"),
        pop = NA_character_,
        pop = case_when(
            species == "ALA" & population %in% c("Broome", "Pilbara", "Exmouth_Gulf") ~ "Coastal_WA",
            species == "ALA" & population == "Shark_Bay" ~ "Shark_Bay",
            species == "ALA" & population %in% c("Ashmore", "Heywood_Shoal", "Scott_Reef") ~ "Offshore_WA",
            species == "ALA" & population == "Gulf_of_Carpentaria" ~ "Gulf_of_Carpentaria",
            species == "ALA" & population == "New_Caledonia" ~ "New_Caledonia",
            .default = pop
        ),
        pop = case_when(
            species == "HMA" & population %in% c("Broome", "Pilbara", "Exmouth_Gulf") ~ "Coastal_WA",
            species == "HMA" & population == "Shark_Bay" ~ "Shark_Bay",
            species == "HMA" & population =="Gulf_of_Carpentaria" ~ "Gulf_of_Carpentaria",
            species == "HMA" & population == "New_Caledonia" ~ "New_Caledonia",
            .default = pop
        ),
        pop = case_when(
            species == "HST" & population %in% c("Broome", "Pilbara", "Exmouth_Gulf", "North_Kimberley") ~ "Coastal_WA",
            species == "HST" & population %in% c("Ashmore", "Heywood_Shoal") ~ "Offshore_WA",
            species == "HST" & population == "Gulf_of_Carpentaria" ~ "Gulf_of_Carpentaria",
            species == "HST" & population == "New_Caledonia" ~ "New_Caledonia",
            .default = pop
        )
    ) |>
    # Generating genome-wide averages
    summarise(
        pixy_pi = sum(count_diffs) / sum(count_comparisons),
        .by = c(species, pop)
    )
```

The π estimates are included in the next table.

## Within population statistics

Below we simply extract the weighted averages for each population. The table
contains four relevant columns - weighted means for nucleotide diversity, 
observed heterozygosity, expected heterozygosity and F~IS~. I've appended the
`pixy` π estimates to compare the impact of calculating nucleotide diversity with
and without invariant data.

```{r within-statistics}
# Number of samples in each subfacet
within <- stats_within |>
    imap(
        \(x, i) {
            x |>
                pluck("weighted.means") |>
                as_tibble() |>
                filter(is.na(weighted_mean_fst)) |>
                select(-c(snp.facet, snp.subfacet, weighted_mean_fst, mean_fst)) |>
                left_join(
                    meta[[i]] |> summarise(n_samples = n(), .by = pop),
                    by = join_by(subfacet == pop)
                )
        }
    ) |>
    list_rbind(names_to = "species") |>
    left_join(pixy_pi, by = join_by(species, subfacet == pop)) |>
    mutate(subfacet = factor(subfacet, levels = pop_levels)) |>
    select(
        Species = species,
        Population = subfacet,
        n_samples,
        `Nuc. diversity (pixy)` = pixy_pi,
        `Nuc. diversity (snpR)` = weighted_mean_pi,
        Ho = weighted_mean_ho, He  = weighted_mean_he,
        Fis = weighted_mean_fis,
        -facet
    ) |>
    arrange(Species, Population)

# Write to file
within |>
    (\(x) split(x, x$Species))() |>
    iwalk(\(x, i) {
        x |>
            write_csv(
                file = here("results", "diversity-statistics", glue::glue("{i}-within-stats.csv"))
            )
    })

# Example of output
within |>
    gt(groupname_col = "Species") |>
    fmt_number(
        columns = 4:8,
        n_sigfig = 2
    ) |>
    as_raw_html()
```

I've left the π calculation based on **just SNPs** in the table. We can see
that the values are much higher than the `pixy` estimations.

# Between population statistics

Lastly we have our F~ST~ measures and Nei's genetic distance measures as heat maps.
We need to manipulate the data a bit as the F~ST~ data don't include the diagonal.

## Clean F~ST~ matrix

Here I add in the missing row/column values from the F~ST~ matrix. We do this
so we have a complete matrix with a diagonal.

```{r}
stats_within |>
    iwalk(\(x, i) {
        mat <- x |>
            pluck("fst.matrix") |>
            pluck("pop") |>
            as.matrix(rownames = "p1")
        
        missing_row <- colnames(mat)[ ! colnames(mat) %in% rownames(mat) ]
        missing_col <- rownames(mat)[ ! rownames(mat) %in% colnames(mat) ]
        
        # Add extra columns
        mat_tmp <- rbind(mat, "tmp" = NA)
        mat_tmp <- cbind("tmp" = NA, mat_tmp)
        
        # Rename to include missing colnames
        colnames(mat_tmp) <- c(missing_col, colnames(mat))
        rownames(mat_tmp) <- c(rownames(mat), missing_row)
        
        # Transpose to lower-tri, then convert to distance matrix, then back to matrix
        mat_tmp <- as.matrix(as.dist(t(mat_tmp), diag = TRUE))
        
        pop_order <- pop_levels[pop_levels %in% colnames(mat_tmp)]
        
        stats_within[[i]]$fst.updated <<- mat_tmp[pop_order, pop_order]
    })
```

## Combine F~ST~ and Nei's genetic distance data into single matrix

```{r get-fst}
df_fst_nei <- map(c("ALA", "HMA", "HST"), \(id){
    
    # Grab Nei genetic distance matrix
    nei <- stats_between[[id]] |> 
        pluck("pop") |> 
        pluck(".base") |> 
        pluck("Nei") |> 
        as.matrix()
    
    # Subset for levels that are present
    pop_order <- pop_levels[pop_levels %in% colnames(nei)]
    
    # Nei - Re-order matrices so the data are West to East
    nei <- nei[pop_order, pop_order]
    nei[upper.tri(nei)] <- NA
    
    # Fst - should already be ordered
    fst <-  stats_within[[id]] |>
        pluck("fst.updated")
    
    # Create new single matrix of all values: Fst upper - Nei lower
    d <- dim(nei)[1]
    new <- matrix(NA, d, d)
    new[lower.tri(new)] <- nei[lower.tri(nei)]
    new[upper.tri(new)] <- fst[upper.tri(fst)]
    colnames(new) <- colnames(fst)
    rownames(new) <- rownames(fst)
    
    # Turn into tibble and pivot_longer and check if values are in
    new_long <- new |>
        as_tibble(rownames = "population") |>
        pivot_longer(names_to = "population2", values_to = "values", cols = 2:(ncol(new) + 1)) |>
        mutate(
            tri_partition = case_when(
                population == population2 ~ "diagonal",
                values %in% nei[lower.tri(nei)] ~ "lower",
                values %in% fst[upper.tri(fst)] ~ "upper"
            ),
            across(starts_with("population"), ~str_replace_all(.x, "_", " ")),
            across(starts_with("population"), ~str_wrap(.x, 8)),
            population = forcats::fct_inorder(population),
            population2 = forcats::fct_inorder(population2)
        )
    
    list(new_long, new)
}) |>
    set_names(c("ALA", "HMA", "HST"))
```

## Write pairwise distance results

First we'll save the combined matrices to file, where upper triangle is F~ST~
and the lower triangle is Nei's genetic distance

```{r}
# Write long form tables used for plotting
df_fst_nei |>
    map(pluck, 1) |>
    iwalk(\(x, i) {
        x |>
            write_csv(
                file = here("results", "diversity-statistics", glue::glue("{i}-between-stats-long.csv"))
            )
    })

# Write matrices
df_fst_nei |> 
    map(pluck, 2) |>
    iwalk(\(x, i) {
        x |>
            as_tibble(rownames = "population") |>
            write_csv(
                file = here("results", "diversity-statistics", glue::glue("{i}-between-stats.csv"))
            )
    })
```

## Pairwise genetic distance heatmaps

Next we'll write a helper function for plotting.

```{r plot-heatmap}
plotHeatMap <- function(
        data,
        species,
        col_fst = c("#dec9e9", "#764E9FFF"), 
        col_nei = c("#d9ed92", "#168aad")
) {
    sp <- case_when(
        species == "ALA" ~ "Aipysurus laevis",
        species == "HMA" ~ "Hydrophis major",
        species == "HST" ~ "Hydrophis stokesii"
    )
    
    data |>
        ggplot() +
        geom_tile(
            data = data |> 
                filter(tri_partition %in%  c("upper", "diagonal")),
            aes(
                x = population, y = population2, 
                fill = values
            ),
            colour = "black"
        ) +
        scale_fill_gradient(
            name = bquote(F[ST]), 
            low = col_fst[[1]], 
            high = col_fst[[2]], 
            na.value = "grey95",
            breaks = seq(0, 0.35, 0.05),
            labels = as.character(seq(0, 0.35, 0.05)),
            limits = c(0, 0.35)
        ) +
        # New scale for bottom triangle
        ggnewscale::new_scale_fill() +
        geom_tile(
            data = data |> 
                filter( tri_partition %in%  c("lower", "diagonal") ),
            aes(x = population, y = population2, fill = values),
            colour = "black"
        ) +
        scale_fill_gradient(
            name = "Nei", 
            low = "#76c893", 
            high = "#1a759f",
            na.value = "grey95",
            breaks = seq(0, 0.03, 0.005),
            labels = as.character(seq(0, 0.03, 0.005)),
            limits = c(0, 0.03)
        ) +
        scale_x_discrete(expand = c(0,0)) +
        scale_y_discrete(expand = c(0,0)) +
        ggtitle(sp) +
        geom_text(
            aes(
                x = population, 
                y = population2, 
                label = signif(values, digits = 2)
            ),
            colour = "white"
        ) +
        theme(
            axis.title = element_blank(),
            axis.text = element_text(size = 14), 
            panel.grid.major = element_blank(),
            plot.title = element_text(face = "italic", hjust = 0.5, size = 16)
        )
}
```

And lastly we'll generate heatmaps of pairwise diversity, where the upper
triangle is F~ST~ and the lower is Nei's genetic distance. These are two different
pairwise measures that should be broadly similar.

```{r}
plot_fst_nei <- df_fst_nei |>
    map(pluck, 1) |>
    imap( \(x, i) { plotHeatMap(x, species = i) } )
design <- "12\n34"

plot_fst_nei <- plot_fst_nei$ALA + plot_fst_nei$HMA + plot_fst_nei$HST + guide_area() +
    plot_layout(
        design = design,
        guides = "collect"
    ) &
    theme(
        legend.box = "hoizontal", 
        legend.position = "bottom",
        legend.key.height = unit(1.5, 'cm'),
        legend.key.width = unit(2.5, "cm"),
        
        legend.key.spacing.y = unit(2, "cm"),
        legend.title = element_text(size = 14, hjust = 0.5),
        legend.text = element_text(size = 14),
        legend.title.position = "top"
    )

ragg::agg_png(
    filename = here("results", "diversity-statistics", "Fst_Nei-heatmap.png"), 
    width = 1500, 
    height = 1200, 
    units = "px", 
    res = 100
)
plot_fst_nei
invisible(dev.off())

plot_fst_nei
```


